---
title: "Human_MI_VanDerWijst"
author: "Carolian Duque"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---

# Overview
the [Single-cell dissection of the immune response after acute myocardial infarction](https://www.ahajournals.org/doi/10.1161/CIRCGEN.123.004374) paper contains:

PBMCs and plasma from 38 patients with a first STEMI at the following time points:  
- time of hospital admission (t0), 
- 24h (t24h) post MI
- 6-8 weeks (t8w) post MI

Their controls are from the LifeLines DEEP cohort which they first excluded any patients with CVD, medication use etc, and then matched by age and sex with their MI patients

used a combination of 10X Genomics v2 and v3 chemistry. Will need to integrate across these to remove batch effects 
they state they use different quality control (QC) thresholds and analyses were performed separately per chemistry. So we will use their QC filtered data (but not normalized)  

# Setup  
Loading Packages 
```{r, message = FALSE, warning=FALSE, results='hide'}
library("Seurat")
library("ScCustomize")
library(dplyr)
library(patchwork)
library(R.utils)
library(ggplot2)
library("HGNChelper")
library("openxlsx")
#library("ggraph")
library("igraph")
#library("tidyverse")
library("data.tree")
library("biomaRt")
library("DESeq2")
```
  
load their QC-ed but not normalized data into seruat. Load metadata
```{r, message=FALSE,warning=FALSE}
stemi_raw <- Read10X("~/van_der_wijst_medRxiv/raw_QC-ed_data/", gene.column = 1, cell.column = 1)
stemi = CreateSeuratObject(counts = stemi_raw)

metadata_github <- read.table("~/metadata.tsv.gz", header = T, row.names = 1, sep = '\t')

stemi <- AddMetaData(stemi, metadata = metadata_github)
rm(stemi_raw)
```

  
quick check to make sure their file is actually filtered by the QC variables
```{r, message = FALSE, warning=FALSE}
stemi[["percent.mt"]] <- PercentageFeatureSet(stemi, pattern = "^MT-",assay="RNA") 
VlnPlot(stemi, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)
VlnPlot(stemi, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0)
VlnPlot(stemi, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0,split.by = "orig.ident",fill.by = "orig.ident",cols = c("red","blue","green","purple"))

```
  
seems like MT is < 15%, nFeatrues is 200 - 8000

# Processing

*quality filter* 

1. remove doublets - They ran scrublet in QC before this 
2. remove high mitochondrial genes. I set my threshold a bit higher since there seem to be a good amount of mitochondrial genes based on the distribution
3. remove genes with really high expression 5000 seems reasonable based on the distribution

```{r}
stemi <- subset(stemi, subset =
  ((orig.ident == "stemi_v3" | orig.ident == "hc_v3") & nFeature_RNA > 200 &  nFeature_RNA < 4500 & percent.mt < 12 & status=="singlet" ) | 
   ((orig.ident == "stemi_v2" | orig.ident == "hc_v2") & nFeature_RNA > 200 &  nFeature_RNA < 3500 & percent.mt < 8 & status=="singlet" ) 
  )

```
  
checking plots again
```{r}
VlnPlot(stemi, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0,split.by = "orig.ident",fill.by = "orig.ident",cols = c("red","blue","green","purple"))
```
  
*Normalize and Scale*  
  
```{r}
stemi <-NormalizeData(stemi, normalization.method = "LogNormalize", scale.factor = 10000) %>% 
  FindVariableFeatures(selection.method = "vst", nfeatures = 3000) 
all.genes <- rownames(stemi)
stemi <- ScaleData(stemi, features = all.genes)  
```
  
*Integrate Datasets*
  
This dataset has evident batch effects. It comes from two different datasets and two different 10x chemistries. Their metadata further indicates lane and other batches. 
  
Make PCA and also do a quick check of batch effects
```{r, warning=FALSE}
stemi <- RunPCA(stemi, verbose = FALSE)
stemi <- RunUMAP(stemi, reduction = "pca", dims = 1:30, verbose = FALSE) %>% 
  FindNeighbors(reduction = "pca", dims = 1:30)

stemi<-FindClusters(
  stemi,
  resolution = 0.4
)

DimPlot(stemi, reduction = "umap",raster = FALSE, group.by = "orig.ident")
DimPlot(stemi, reduction = "umap",raster = FALSE, split.by = "orig.ident")
```
  
v2 vs v3 chemistry seems to have a huge effect. and their paper performed integration based on chemistry and dataset, so we will do the same here.

We have done harmony for others, so we will do the same here. 

```{r}
library("harmony")
stemi <- RunHarmony(stemi, "orig.ident")
```
  
# Clustering
  
Check elbow plot here
```{r}
ElbowPlot(stemi)
```
  
UMAPS. 

check harmony worked
```{r, message = FALSE}
stemi <- RunUMAP(stemi, reduction = "harmony", dims = 1:30, verbose = FALSE) %>% FindNeighbors(reduction = "harmony", dims = 1:30)

stemi<-FindClusters(
  stemi,
  resolution = 0.5
)

DimPlot(stemi, reduction = "umap",raster = FALSE, group.by = "orig.ident")
DimPlot(stemi, reduction = "umap",raster = FALSE, split.by = "orig.ident")

```
# scType

Load necessary functions for scType
```{r}
#scType gene set preparation function and annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R"); source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```
  
We used an edited cell marker database
* tissueType - tissue type
* cellName - cell type
* geneSymbolmore1 - positive marker genes
* geneSymbolmore2 - marker genes not expected to be expressed by a cell type)  

```{r}
DB = "ScTypeDB_Carolina_edited.xlsx"

Immune_gs_list = gene_sets_prepare(DB, c("Immune system") )

#create cell-type by cell matrix
es.max = sctype_score(scRNAseqData = stemi[["RNA"]]@scale.data, scaled = TRUE, 
                      gs = Immune_gs_list$gs_positive, gs2 = Immune_gs_list$gs_negative) 

# merge by cluster
cL_resutls = do.call("rbind", 
                     lapply(unique(stemi@meta.data$seurat_clusters), function(cl){
    es.max.cl = sort(rowSums(es.max[,rownames(stemi@meta.data[stemi@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(stemi@meta.data$seurat_clusters==cl)), 10)
}))

sctype_scores = cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] = "Unknown"
print(sctype_scores[,1:3])


stemi@meta.data$scType_Immune = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  stemi@meta.data$scType_Immune[stemi@meta.data$seurat_clusters == j] = as.character(cl_type$type[1])
}

stemi@meta.data$condition = ""
stemi@meta.data = stemi@meta.data %>% mutate(condition = case_when(timepoint.final == "Baseline" ~ "0h_postMI",
                                                                   timepoint.final == "t24h" ~ "24h_postMI",
                                                                   timepoint.final == "t8w" ~ "8w_postMI",
                                                                   timepoint.final == "UT" ~ "Healthy_Control"))

Idents(object = stemi) <- "scType_Immune"
```
simplified annotations
```{r}
stemi@meta.data$scType_Immune_edited = stemi@meta.data$scType_Immune

stemi@meta.data=stemi@meta.data %>%
  mutate(scType_Immune_edited = case_when(
    scType_Immune_edited=="Plasma B cells"|scType_Immune_edited=="Naive B cells"|scType_Immune_edited=="Pre-B cells"|scType_Immune_edited=="Memory B cells" ~ "B_Cells",
    TRUE ~ "Other Immune Cells"))
```

```{r}

DimPlot(stemi, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune_edited', split.by = "condition",raster=FALSE)
ggsave("~/UMAP_wijst_blood_scType_disease_simplified_B_0_24_hc_bysex.svg",width = 12.5,height=10.5)
```

create an object with 24h and another with 8w
```{r}
stemi_24h = subset(x = stemi, subset = condition != "8w_postMI" )
stemi_8w = subset(x = stemi, subset = condition != "24h_postMI" )
```

# DGE

use AggregateExpression() to sum together gene counts of all the cells from the same sample for each cell type. This results in one gene expression profile per sample and cell type. We can then perform DE analysis using DESeq2 on the sample level. This treats the samples, rather than the individual cells, as independent observations.
  
aggregate expression and clean up metadata
```{r, message=FALSE}
pseudo_stemi <- AggregateExpression(stemi, assays = "RNA",slot ="counts", return.seurat = T, group.by = c("condition", "assignment.final", "scType_Immune_edited"))

# gather metadata 
pseudo_stemi_metadata=pseudo_stemi@meta.data

pseudo_stemi_metadata$condition  <- sapply(strsplit(Cells(pseudo_stemi), split = "_"), function(x) paste(x[1:2], collapse = "_"))
pseudo_stemi_metadata$assignment.final <- sapply(strsplit(Cells(pseudo_stemi), split = "_"), "[", 3)
pseudo_stemi_metadata$scType_Immune_edited <- sapply(strsplit(Cells(pseudo_stemi), split = "_"), function(x) paste(x[4:length(x)], collapse = "_"))

pseudo_stemi_metadata$celltype.cond <- paste(pseudo_stemi_metadata$scType_Immune_edited, pseudo_stemi_metadata$condition, sep = "_")



temp_metadata=stemi@meta.data
temp_metadata=subset(temp_metadata, select=c("assignment.final","condition","gender","age","chem","batch"))
temp_metadata=unique(temp_metadata)

pseudo_stemi_metadata$ROWNAMES=rownames(pseudo_stemi_metadata)
pseudo_stemi_metadata$row_order  <- 1:nrow(pseudo_stemi_metadata)

pseudo_stemi_metadata = merge(pseudo_stemi_metadata,temp_metadata,by=c("assignment.final","condition"),all.x=TRUE)
rownames(pseudo_stemi_metadata)=pseudo_stemi_metadata$ROWNAMES
pseudo_stemi_metadata =pseudo_stemi_metadata[order(pseudo_stemi_metadata$row_order), ]

pseudo_stemi_metadata$celltype.cond.sex <- paste(pseudo_stemi_metadata$celltype.cond, pseudo_stemi_metadata$gender, sep = "_")

# add back the metadata to the seurat object
pseudo_stemi=AddMetaData(pseudo_stemi,metadata = pseudo_stemi_metadata)

# subset to only B and T cells to make the sparse matrix smaller
pseudo_stemi=subset(pseudo_stemi,scType_Immune_edited=="B_Cells"|scType_Immune_edited=="CD4_T_Cells")

pseudo_stemi_metadata=pseudo_stemi@meta.data
```
    
## DESeq2 not sex stratified 
   
I am controlling for the 10x chemistry that had a huge batch effect earlier. You can see it again in the PCA plots below.  
  
I also control for gender since there are usually big gender based differences

For the condition of interest you can see it start to separate on PC5. 

```{r, message=FALSE, warning=FALSE}
library(GGally)
dds_pseudo_stemi=DESeqDataSetFromMatrix(pseudo_stemi@assays$RNA@counts, 
                              colData = pseudo_stemi_metadata, 
                              design = ~ gender + chem + celltype.cond)

vst_pseudo_stemi=vst(dds_pseudo_stemi)
rv  <- rowVars(assay(vst_pseudo_stemi),useNames=TRUE)
pcagenes  <-  order(rv , decreasing=TRUE)[seq_len(min(500, length(rv )))]
pca  <- prcomp(t(assay(vst_pseudo_stemi)[pcagenes ,]))
percentVar  <- pca$sdev^2 / sum( pca$sdev^2 )

DESeq2::plotPCA(vst_pseudo_stemi, intgroup = "scType_Immune_edited")
DESeq2::plotPCA(vst_pseudo_stemi, intgroup = "chem")

ggpairs(pca$x, columns = 1:5, aes(colour=pseudo_stemi_metadata$gender), upper =list(continuous= "blank"),diag = list(continuous = wrap("densityDiag",alpha=0.5)),legend=c(2,1),lower = list(continuous = wrap("points", alpha = 0.9, size=0.7)))
  theme_bw() + labs(colour="gender")
  
ggpairs(pca$x, columns = 1:5, aes(colour=pseudo_stemi_metadata$condition), upper =list(continuous= "blank"),diag = list(continuous = wrap("densityDiag",alpha=0.5)),legend=c(2,1),lower = list(continuous = wrap("points", alpha = 0.9, size=0.7)))
  theme_bw() + labs(colour="condition")
```
### 8w vs HC or 0h. 
  
```{r}

dds_pseudo_stemi <- DESeq(dds_pseudo_stemi)

res_8_v_hc_bcells <- results(dds_pseudo_stemi, 
               contrast = c("celltype.cond","B_Cells_8w_postMI","B_Cells_Healthy_Control"),
               alpha = 0.05)
summary(res_8_v_hc_bcells)
sig_res_8_v_hc_bcells = subset(as.data.frame(res_8_v_hc_bcells), padj < 0.05& (log2FoldChange > 0.58 | log2FoldChange< -0.58 ))
write.csv(as.data.frame(res_8_v_hc_bcells),"van_der_wijst_medRxiv/DESeq_res_8_v_hc_bcells.csv")
```


### 24h vs HC or 0h. 
  
```{r}

res_24_v_hc_bcells <- results(dds_pseudo_stemi, 
               contrast = c("celltype.cond","B_Cells_24h_postMI","B_Cells_Healthy_Control"),
               alpha = 0.05)
summary(res_24_v_hc_bcells)

sig_res_24_v_hc_bcells = subset(as.data.frame(res_24_v_hc_bcells), padj < 0.05& (log2FoldChange > 0.524 | log2FoldChange< -0.524 ))
write.csv(as.data.frame(res_24_v_hc_bcells),"DESeq_res_24_v_hc_bcells.csv")
```

