---
title: "Human_MI_reanalysis_Rao"
author: "Carolian Duque"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview
the [Resolving the intertwining of inflammation and fibrosis in human heart failure at single-cell level](https://link.springer.com/article/10.1007/s00395-021-00897-1#Sec20) paper contains

LV - left ventricle 
RV - right ventricle 
P - CD45+
N - CD45-
Bld - blood

for ICM patients: 
MI - LV infarcted miocardium
NMI - LV non-infarced miocardium 

** They should have 2 normal samples and 3 dcm samples, but they seem to only have normal_1 up and DCM_2 and DCM_3 up on GEO

RM sampels are CD3+ so only T cells. They used for TCR sequencing
The rest are bulk RNA seq


# Setup
Loading Packages
```{r}
library("Seurat")
library("ScCustomize")
library(dplyr)
library(patchwork)
library(R.utils)
library(ggplot2)
library("HGNChelper")
library("openxlsx")
#library("ggraph")
library("igraph")
#library("tidyverse")
library("data.tree")
library("biomaRt")
library("cli")
library("pbapply")
library("Matrix")
```


read in their CellRanger? data
```{r}
data_dir <- "raw/cellranger"
list_dir <- list.files(data_dir,pattern =  "*barcodes.tsv.gz")
sample_ID <- gsub(".*_(.+)_.*", "\\1", list_dir)
GSE145154_10X <- Read10X_GEO(data_dir =data_dir)
# this makes a list of matrices (single modality data)
SeuratSet_GEO <- lapply(GSE145154_10X, CreateSeuratObject)
GEO_MI_dataset <- merge(SeuratSet_GEO[[1]], y = SeuratSet_GEO[2:length(SeuratSet_GEO)], add.cell.ids = sample_ID, project = "GEO_MI")
```

Extract metadata
```{r}
metadata_GEO_MI_dataset <- GEO_MI_dataset@meta.data
```

update metadata with clinical data as per in overview informaton
```{r}
metadata_GEO_MI_dataset$tissue = ""
metadata_GEO_MI_dataset$condition = ""
metadata_GEO_MI_dataset$disease = ""
metadata_GEO_MI_dataset$Enrichment = ""

metadata_GEO_MI_dataset$temp= sub("_.*", "", rownames(metadata_GEO_MI_dataset))

metadata_GEO_MI_dataset=metadata_GEO_MI_dataset %>%
  mutate(disease = case_when(
    startsWith(temp, "N") ~ "Normal",
    startsWith(temp, "DCM") ~ "DCM",
    startsWith(temp, "ICM") ~ "ICM"
    ))

metadata_GEO_MI_dataset=metadata_GEO_MI_dataset %>%
  mutate(Enrichment = case_when(
    endsWith(temp, "N") ~ "CD45-",
    endsWith(temp, "P") ~ "CD45+",
    grepl("Bld", temp) ~ "CD45+"
    ))

metadata_GEO_MI_dataset=metadata_GEO_MI_dataset %>%
  mutate(tissue = case_when(
    grepl("LV", temp) ~ "LV",
    grepl("RV", temp) ~ "RV",
    grepl("MI", temp) ~ "LV",
    grepl("Bld", temp) ~ "Blood"
  ))

metadata_GEO_MI_dataset=metadata_GEO_MI_dataset %>%
  mutate(condition = case_when(
    grepl("-MI", temp) ~ "Infarcted",
    grepl("-NMI", temp) ~ "Non-infarcted",
    (disease=="ICM" & grepl("LV", temp)) ~ "Infarcted",
    (disease=="ICM" & grepl("RV", temp)) ~ "Non-infarcted",
    TRUE ~ "Not-ICM"
  ))

GEO_MI_dataset <- AddMetaData(GEO_MI_dataset, metadata = metadata_GEO_MI_dataset)
```

subset database to ICM and Normal blood
```{r}
sub_blood_MI = subset(x = GEO_MI_dataset, subset = tissue == "Blood" & disease != "DCM" )

metadata_sub_blood_MI = sub_blood_MI@meta.data
```

# Process the data sets
*1. Quality check*  
* very low # of unique genes or total # of molecules detected can represent low quality cells or empty droplets
* very high # of unique genes or total # of molecules detected can represent doublets (hastag demultiplexing helps with this)
* very high # of mitochondrial genes can indicate dying or low quality cells  
```{r}

sub_blood_MI[["percent.mt"]] <- PercentageFeatureSet(sub_blood_MI, pattern = "^MT-",assay="RNA")
VlnPlot(sub_blood_MI, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

plot1 <- FeatureScatter(sub_blood_MI, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(sub_blood_MI, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2
```
*filter out very low and very high expression and very high mitochondrial genes:*  
```{r}


sub_blood_MI.sce <- as.SingleCellExperiment(sub_blood_MI)
sub_blood_MI.sce <- scDblFinder(sub_blood_MI.sce, samples="temp")
sub_blood_MI = as.Seurat(sub_blood_MI.sce)
test=sub_blood_MI@meta.data
sub_blood_MI <- subset(sub_blood_MI, subset = nFeature_RNA > 200 & percent.mt < 5 & scDblFinder.class=="singlet")

```

*Normalize and Scale*  
Since we did the regular seurat normalization strategy for the mouse data, we will do the same here. 

```{r}
sub_blood_MI <-NormalizeData(sub_blood_MI, normalization.method = "LogNormalize", scale.factor = 10000) %>% 
  FindVariableFeatures(selection.method = "vst", nfeatures = 3000) 
all.genes <- rownames(sub_blood_MI)
sub_blood_MI <- ScaleData(sub_blood_MI, features = all.genes)  
```

*Integrate Datasets*
This dataset has evident batch effects, and appears to be by sample (at least visually). Since I don't know their batches, I will use sample ID as the batch. 

Harmony needs a pca
```{r}
sub_blood_MI <- RunPCA(sub_blood_MI, verbose = FALSE)
ElbowPlot(sub_blood_MI)
```

```{r}
library("harmony")
sub_blood_MI <- RunHarmony(sub_blood_MI, "temp")
```


# Clustering


## UMAP
resolution parameter sets the ‘granularity’ of the downstream clustering, with increased values leading to a greater number of clusters. Generally resultion is 0.4-1.2, and optimal resolution often increases for larger datasets
```{r}
sub_blood_MI <- RunUMAP(sub_blood_MI, reduction = "harmony", dims = 1:30, verbose = FALSE) %>% FindNeighbors(reduction = "harmony", dims = 1:30)

sub_blood_MI<-FindClusters(
  sub_blood_MI,
  resolution = 0.5
)
```

# scType

Load necessary functions for scType
```{r}
#scType gene set preparation function and annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R"); source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```


We used an edited cell marker database
* tissueType - tissue type
* cellName - cell type
* geneSymbolmore1 - positive marker genes
* geneSymbolmore2 - marker genes not expected to be expressed by a cell type)  

```{r}
DB = "ScTypeDB_Carolina_edited.xlsx"

Immune_gs_list = gene_sets_prepare(DB, c("Immune system") )

#create cell-type by cell matrix
es.max = sctype_score(scRNAseqData = sub_blood_MI[["RNA"]]@scale.data, scaled = TRUE, 
                      gs = Immune_gs_list$gs_positive, gs2 = Immune_gs_list$gs_negative) 

# merge by cluster
cL_resutls = do.call("rbind", 
                     lapply(unique(sub_blood_MI@meta.data$seurat_clusters), function(cl){
    es.max.cl = sort(rowSums(es.max[,rownames(sub_blood_MI@meta.data[sub_blood_MI@meta.data$seurat_clusters==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(sub_blood_MI@meta.data$seurat_clusters==cl)), 10)
}))

sctype_scores = cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/4] = "Unknown"
print(sctype_scores[,1:3])

sub_blood_MI@meta.data$scType_Immune = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  sub_blood_MI@meta.data$scType_Immune[sub_blood_MI@meta.data$seurat_clusters == j] = as.character(cl_type$type[1])
}

Idents(object = sub_blood_MI) <- "scType_Immune"

DimPlot(sub_blood_MI, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune')

DimPlot(sub_blood_MI, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune', split.by = "disease")

```
make simplified version of the UMAP with only the B cells shown and everything else in grey: 

```{r}
sub_blood_MI@meta.data$scType_Immune_edited = sub_blood_MI@meta.data$scType_Immune

sub_blood_MI@meta.data=sub_blood_MI@meta.data %>%
  mutate(scType_Immune_edited = case_when(
    scType_Immune_edited=="Plasma B cells"|scType_Immune_edited=="Naive B cells"|scType_Immune_edited=="Pre-B cells" ~ "B Cells",
    scType_Immune_edited!="B Cells" ~ "Other Immune Cells"
  ))

Idents(object = sub_blood_MI) <- "scType_Immune_edited"
DimPlot(sub_blood_MI, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune_edited', split.by = "disease", cols = c("#ff6666","grey"))
ggsave("UMAP_blood_scType_disease_simplified.svg",width = 12.5,height=6.5)
```


