---
title: "postMI_allcells"
author: "Carolina Duque"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
---


#Prep

Loading Packages
```{r}
library("Seurat")
library(dplyr)
library(patchwork)
library(R.utils)
library(ggplot2)
library("HGNChelper")
library("openxlsx")
library("igraph")
library("data.tree")
library("biomaRt")
```



# Setup   
 
Loading h5 file from cellranger output.
```{r,message = FALSE, warning=FALSE}
set.seed(1234)
setwd(dir="Jana_MI_allCells_round2")
h5_files <- list.files(pattern =  "*filtered_feature_bc_matrix.h5", recursive = TRUE, full.names = TRUE)
list_h5_data <- lapply(h5_files, Read10X_h5)

names(list_h5_data) <- sub("^.*/(.*)_final/per.*", "\\1", h5_files)


h5_seurat_list <- lapply(X=list_h5_data, FUN=function(x) {CreateSeuratObject(counts = x$`Gene Expression`)})
all_cells_seurat  <- merge(h5_seurat_list[[1]], y = h5_seurat_list[2:length(h5_seurat_list)], 
                  add.cell.ids = c("hash1","hash2","hash6"), project = "jana_Bcell_MI_mice_2")
```

create metadata
```{r}
all_cells_metadata=all_cells_seurat@meta.data

all_cells_metadata$hashtag=gsub("^(.*)_.*$", "\\1", rownames(all_cells_metadata))

all_cells_metadata <- all_cells_metadata %>%
  mutate(condition = case_when(hashtag=="hash1" ~ "WT_MI",
                               hashtag=="hash2" ~ "MHC_KO_MI",
                               hashtag=="hash6" ~ "WT_sham"))

all_cells_seurat <- AddMetaData(all_cells_seurat, metadata = all_cells_metadata)

```



# Process the data sets 
  
*1. Quality check*   
  
```{r}

Idents(all_cells_seurat)=all_cells_seurat$condition
all_cells_seurat[["percent.mt"]] <- PercentageFeatureSet(all_cells_seurat, pattern = "^mt-",assay="RNA") 

VlnPlot(all_cells_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3)

plot1 <- FeatureScatter(all_cells_seurat, feature1 = "nCount_RNA", feature2 = "percent.mt")
plot2 <- FeatureScatter(all_cells_seurat, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
plot1 + plot2

VlnPlot(all_cells_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0,split.by = "condition")

```  

doublet finder. doublet rate for new GEM-X is 0.04% per 1000 cells. Looking at the assignment_confidence_table there were 43152 cells recovered of which 4146 were unassigned / blank and 7190 were doublets (16.6%). Based on 43152 cells recovered we would expect 17.2% doublets. hashtag would presumably find 1/2 of the doublets given that there are 3 hashtags in this case (ie it found 1-2,1-6,2-6 doublets but not 1-1, 2-2, 6-6)). i think to be conservative i will say an expected doublet rate of 17.2% /2 = 8.6%

```{r}
all_cells_seurat <- JoinLayers(all_cells_seurat)

library(scDblFinder)
set.seed(12345)
all_cells_seurat.sce <- as.SingleCellExperiment(all_cells_seurat)
set.seed(12345)
all_cells_seurat.sce <- scDblFinder(all_cells_seurat.sce, dbr = 0.086)
all_cells_seurat = as.Seurat(all_cells_seurat.sce, data = NULL)
all_cells_metadata=all_cells_seurat@meta.data
```


*2. filter out very low and very high expression and very high mitochondrial genes:*  
  
Based on the distribution of your data I am going to filter out mitochondrial genes >10%, and unique features <200  and > 4800


```{r}
all_cells_seurat <- subset(all_cells_seurat, subset = nFeature_RNA > 200 & nFeature_RNA < 4800 & percent.mt < 10 & scDblFinder.class =="singlet")

VlnPlot(all_cells_seurat, features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), ncol = 3, pt.size=0,split.by = "condition")
```

```{r}
table(all_cells_seurat$condition)
```


*3. Normalize and Scale*  
```{r}
all_cells_seurat <-NormalizeData(all_cells_seurat, normalization.method = "LogNormalize", scale.factor = 10000) %>% 
  FindVariableFeatures(selection.method = "vst", nfeatures = 3000) 
all.genes <- rownames(all_cells_seurat)
all_cells_seurat <- ScaleData(all_cells_seurat, features = all.genes)  
```

there is just one pool so no need to remove batch effects

```{r}
set.seed(12345)
all_cells_seurat <- RunPCA(all_cells_seurat, verbose = FALSE) %>% 
  RunUMAP(reduction = "pca", dims = 1:30, verbose = FALSE) %>% 
  FindNeighbors(reduction = "pca", dims = 1:30)

set.seed(12345)
all_cells_seurat<-FindClusters(
  all_cells_seurat,
  resolution = 0.5
)

DimPlot(all_cells_seurat, reduction = "umap",raster = FALSE, group.by = "condition")
```


# scType

Load necessary functions for scType
```{r}
#scType gene set preparation function and annotation function
source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/gene_sets_prepare.R"); source("https://raw.githubusercontent.com/IanevskiAleksandr/sc-type/master/R/sctype_score_.R")
```

since these are mice, we need to convert to human gene symbols for scType to work well.

```{r}
mouse = data.frame("names"=rownames(all_cells_seurat))
mouse = data.frame(mouse$names[1:10])
require("biomaRt")

humanMart = useMart("ENSEMBL_MART_ENSEMBL", dataset = "hsapiens_gene_ensembl", host="https://jul2023.archive.ensembl.org")
mouseMart = useMart("ENSEMBL_MART_ENSEMBL", dataset = "mmusculus_gene_ensembl", host ="https://jul2023.archive.ensembl.org")

#getLDS is not working. might be a connection issue, so i just grabbed it from the biomart website. upload below
human_mouse=read.csv("mouse_human_symbol_mart_export.txt")
genesV2=human_mouse
genesV2 = genesV2 %>%
  dplyr::rename("HGNC.symbol"="Human.gene.name",
         "MGI.symbol"="Gene.name")
genesV2$Human.gene.stable.ID <- NULL

humanx <- genesV2[!duplicated(genesV2$MGI.symbol), ]


library("tibble")

human=data.frame("MGI.symbol"=rownames(all_cells_seurat))
human$order = 1:nrow(human)

human =  human %>% 
  left_join(humanx, by="MGI.symbol")

human =  human %>%
  mutate(HGNC.symbol_edit= ifelse(HGNC.symbol=="",MGI.symbol,HGNC.symbol))

human =human[order(human$order), ]

  
rownames(all_cells_seurat) <- human$HGNC.symbol_edit
```


Will use our own T cell marker data since scType markers are designed for human. This data was based on the expression markers from the following papers and from cellmarker 2.0

- see B cell and T cell analyses for those papers
- https://www.cell.com/iscience/fulltext/S2589-0042(23)01836-9
- https://www.ahajournals.org/doi/full/10.1161/CIRCULATIONAHA.119.045115
- [Kimmel J, et al (2019) Genome Res.](https://genome.cshlp.org/content/29/12/2088)
- Forte et all.  https://www.cell.com/cell-reports/fulltext/S2211-1247(20)30162-5?_returnURL=https%3A%2F%2Flinkinghub.elsevier.com%2Fretrieve%2Fpii%2FS2211124720301625%3Fshowall%3Dtrue 
- [cellmarker 2.0](http://bio-bigdata.hrbmu.edu.cn/CellMarker/)


```{r}
set.seed(12345)
DB = "scType_CD_edited_heart_immune.xlsx"

Immune_gs_list = gene_sets_prepare(DB, c("Immune system") )

#create cell-type by cell matrix
es.max = sctype_score(scRNAseqData = GetAssayData(all_cells_seurat, slot = "scale.data", assay = "RNA"), scaled = TRUE, 
                      gs = Immune_gs_list$gs_positive, gs2 = Immune_gs_list$gs_negative) 

# merge by cluster
cL_resutls = do.call("rbind", 
                     lapply(unique(all_cells_seurat@meta.data$seurat_cluster), function(cl){
    es.max.cl = sort(rowSums(es.max[,rownames(all_cells_seurat@meta.data[all_cells_seurat@meta.data$seurat_cluster==cl, ])]), decreasing = !0)
    head(data.frame(cluster = cl, type = names(es.max.cl), scores = es.max.cl, ncells = sum(all_cells_seurat@meta.data$seurat_cluster==cl)), 10)
}))

sctype_scores = cL_resutls %>% group_by(cluster) %>% top_n(n = 1, wt = scores)  

# set low-confident (low ScType score) clusters to "unknown"
sctype_scores$type[as.numeric(as.character(sctype_scores$scores)) < sctype_scores$ncells/5] = "Unknown"
print(sctype_scores[,1:3])


all_cells_seurat@meta.data$scType_Immune = ""
for(j in unique(sctype_scores$cluster)){
  cl_type = sctype_scores[sctype_scores$cluster==j,]; 
  all_cells_seurat@meta.data$scType_Immune[all_cells_seurat@meta.data$seurat_cluster == j] = as.character(cl_type$type[1])
}

Idents(object = all_cells_seurat) <- "scType_Immune"
```

```{r}
Idents(object = all_cells_seurat) <- "scType_Immune"
DimPlot(all_cells_seurat, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune', raster = FALSE,label.size = 2.5)

```


```{r,  fig.width=15, fig.height =6 }
Idents(object = all_cells_seurat) <- "scType_Immune"
DimPlot(all_cells_seurat, reduction = "umap", label = TRUE, repel = TRUE, group.by = 'scType_Immune', split.by = "condition",raster = FALSE,label.size = 4.5) + ggtitle("scType_annotations") + theme(legend.text = element_text(size = 10))
ggsave("UMAP_scType_Jana_MI_allcells_round2.pdf",width = 15,height=6)
```


# DGE

switch back to mouse genes
```{r}
rownames(all_cells_seurat@assays$RNA) <- human$MGI.symbol
```


so we don't have biological replicates, so wouldn't be possible to aggregate expression and then run DESeq since it would involve comparing 1 sample vs 1 sample

```{r, message=FALSE, warning=FALSE, eval=FALSE}
library(GGally)
library(DESeq2)
library(ggfortify)
library(plotly)
#library(glmGamPoi)
library(BiocParallel)

all_cells_seurat$condition = factor(all_cells_seurat$condition, levels = c("WT_sham","WT_MI","MHC_KO_MI"))

all_cells_seurat@meta.data=all_cells_seurat@meta.data %>%
  mutate(celltype.cond = paste(scType_Immune_edited,condition, sep="_"))
all_cells_seurat@meta.data=all_cells_seurat@meta.data %>%
  mutate(specificcelltype.cond = paste(scType_Immune,condition, sep="_"))
```

running into this error so need to adjust count matrix as in the chunk bellow
Error in estimateSizeFactorsForMatrix(counts(object), locfunc = locfunc,  : 
  every gene contains at least one zero, cannot compute log geometric means
```{r, eval=FALSE}
genes_to_keep <- rowSums(all_cells_seurat@assays$RNA$counts > 1) >= 10 # genes expressed (count > 1) in at least 10 cells
features_keep=rownames(all_cells_seurat)[genes_to_keep]

all_cells_seurat_deseq=all_cells_seurat
all_cells_seurat_deseq <- subset(all_cells_seurat, features = features_keep)

all_cells_seurat_deseq <- subset(all_cells_seurat_deseq, subset = scType_Immune_edited!="Unknown")

all_cells_seurat_deseq@assays$RNA@counts<-as.matrix(all_cells_seurat_deseq@assays$RNA@counts)+1
```


for summarized cell types

```{r, eval=FALSE}
summarized_cell_types = unique(all_cells_seurat@meta.data$scType_Immune_edited)
summarized_cell_types = setdiff(summarized_cell_types,"Unknown")
summarized_cell_types_seurat <- list()



# Loop over summarized cell types and create subsets
for (x in summarized_cell_types) {
  summarized_cell_types_seurat[[paste(x, "seurat", sep = "_")]] <- subset(all_cells_seurat_deseq, subset = scType_Immune_edited == x)
}




dds_summarized_list <- list()

for (x in summarized_cell_types) {
 seurat_name <- paste(x, "seurat", sep = "_")
 seurat_obj <- summarized_cell_types_seurat[[seurat_name]]
 
 seurat_obj$celltype.cond = factor(seurat_obj$celltype.cond, levels = c(paste(x, "WT_sham", sep = "_"),paste(x, "WT_MI", sep = "_"),paste(x, "MHC_KO_MI", sep = "_")))
set.seed(12345)
dds_summarized_list[[paste(x, "dds", sep = "_")]] <- DESeqDataSetFromMatrix(seurat_obj@assays$RNA$counts,
                              colData = seurat_obj@meta.data, 
                              design = ~ celltype.cond)
}

for (x in seq_along(dds_summarized_list)) {
  
  set.seed(12345)
  dds_summarized_list[[x]] <- DESeq(
    dds_summarized_list[[x]], 
    useT = TRUE, 
    minmu = 1e-6, 
    minReplicatesForReplace = Inf
  )
}


summarized_ddsResults <- list()
for (x in summarized_cell_types) {
  dds_name <- paste(x, "dds", sep = "_")
 dds_obj <- dds_summarized_list[[dds_name]]
 
summarized_ddsResults[[paste(x,"WT_MI_v_Sham",sep="_")]] <- results(dds_obj, 
                               contrast = c("celltype.cond",paste(x,"WT_MI",sep="_"),paste(x,"WT_sham",sep="_")))

summarized_ddsResults[[paste(x,"KO_MI_v_Sham",sep="_")]] <- results(dds_obj, 
                               contrast = c("celltype.cond",paste(x,"MHC_KO_MI",sep="_"),paste(x,"WT_sham",sep="_")))

summarized_ddsResults[[paste(x,"KO_MI_v_WT_MI",sep="_")]] <- results(dds_obj, 
                               contrast = c("celltype.cond",paste(x,"MHC_KO_MI",sep="_"),paste(x,"WT_MI",sep="_")))
}


names(summarized_ddsResults) <- gsub("/", "_", names(summarized_ddsResults))
names(summarized_ddsResults) <- gsub(" ", "_", names(summarized_ddsResults))

for (i in seq_along(summarized_ddsResults)) {
  name_result <- names(summarized_ddsResults)[i]
  write.csv(as.data.frame(summarized_ddsResults[i]),paste0("DEG/",name_result,".csv"))
}
```
